---
title: "Lecture 2. Assumptions and Interest Rate Mechanics"
output: html_document
---

<style type="text/css">

body{ /* Normal  */
      font-size: 18px;
      font-family: "Times New Roman", Times, serif;
  }
td {  /* Table  */
  font-size: 20px;
}
h1.title {
  font-size: 24px;
  color: DarkBlue;
}
h1 { /* Header 1 */
  font-size: 22px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 22px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 20px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 20px;
}
</style>

# Purpose

* To develop valuation and risk measurement principles, we need assumptions about investor preferences, markets, and interest rate mechanics.
  + (Required) No costless arbitrage opportunities
  + (Optional) Frictionless markets
  + (Optional) Continuously compounded interest rates


# Law of one Price
* No costless arbitrage opportunities
  + Suppose asset trades at two locations. If prices are not identical, individuals will buy cheapest and sell most expensive until price converges.
  + IBM trades in multiple locations. Suppose IBM last traded at 120.75 on NYSE and at 120.25 on PSE. 
    - Buy IBM at 120.25 on PSE and sell at 120.75 on NYSE.
    - Costless arbitrage profit is 0.50.
    - Risk-free arbitrage profit can be earned. 	
* Arbitrage has no risk. Buy one asset and sell perfect substitute.
* Costless arbitrage is arbitrage without investment. 


# Frictionless Markets - Assumptions for convenience
* Markets are assumed to be frictionless. 
  + Assumption of convenience.
  + Permits clearer focus on economics of valuation and risk measurement. 
* Frictionless markets
  + No bid-ask spreads or commissions
  + No taxes
  + Unlimited borrowing or lending at the same risk-free rate
  + Freedom to short
  + Can trade at any time and in any quantity


# Transaction costs and the bid-ask spread
* Buying and selling a financial asset 
  + Brokers : commissions
  + Market-makers : bid-ask spread
* Example : Buy and sell 100 shares of XYZ
  + XYZ : bid = \$49.75 , ask = \$50, commission = \$15
  + Buy : (100 $\times$ \$50) + \$ 15 = \$5,015
  + Sell : (100 $\times$ \$ 49.75) + \$ 15 = \$4,960
  + Transaction cost : \$5,015 - \$ 4,960 = \$ 55


# Short-selling
* When price of an asset is expected to fall
  + First : borrow and sell an asset (get \$\$)
  + Then : buy back and return the asset (pay \$)
  + If price fell in the mean time : Profit = \$\$ - \$ 
  + The lender must be compensated for dividends received (lease-rate)
  + Example: short-sell IBM stock for 90 days


# Short-selling
* Why short-sell?
  + Speculation
  + Financing
  + Hedging
* Interest received from lender on collateral
  + Scarcity decreases the interest rate
  + Repo rate in bond markets
  + Short rebate in the stock market
* We ignore these costs for convenience


# Frictionless Markets
* No bid-ask spreads or commissions.
  + As long as they are known, effects of trading costs can be modeled.
  + E.g., to determine whether an arbitrage profit exists in the IBM example, compare ask price of IBM on PSE with bid price on NYSE.


# Frictionless Markets
* No taxes.
  + Reduces gains. No effect on our results
  + May affect different forms of income differently. 
    - Capital gains vs. ordinary income
    - As long as different tax rates are known, effects can be modeled. 
* Unlimited borrowing or lending at the risk-free rate of interest. 
  + Borrowing rate is higher than lending rate.
  + As long as rates are known, effects can be modeled. 


# Frictionless Markets
* Freedom to short, with full use of proceeds.
  + Assumption is true for large investors in most financial markets.
  + If use of short sales proceeds is prohibited, effects can be modeled (as an opportunity cost of funds).
* Can trade at any time and in any quantity.
  + With deep and liquid markets, trades can be executed "in size" without affecting price.
  + With less liquid markets, large trades may move prices.
  + As long as price impact can be modeled, effects on valuation and risk measurement can be modeled. 


# Valuation by replication
* Arbitrage : Situation in which one can generate positive cash flow by simultaneously buying and selling related assets, with no net investment and with no risk (free money)
* No costless arbitrage opportunities together with frictionless markets permits valuation-by replication.
* If two assets or portfolios of assets have same future payoffs, they must have same current price


# Two assets, A and B, trade in market. Below are their current and future values. 


``` {r abTable, echo=FALSE, message=FALSE, results='asis', comment=""}
if(!require(pander)){install.packages("pander");library(pander)}
ab_table = read.csv("./data/ab_table.csv")
names(ab_table) = c("","Time","")
pandoc.table(ab_table)
``` 


# Illustration
* New asset, C, is introduced into marketplace and has terminal value of 20. What is its current price? 
\end{itemize}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcc}
\hline
   &  \multicolumn{2}{c}{Time} \\
\cline{2-3}
Asset   & 0     & 1     \\
\hline \\
A & 5     & 10      \\
B & 15    & 30    \\
C & 12     & 20 \\


# Illustration : Solution
* Buying $\frac{1}{2}$ unit of A and $\frac{1}{2}$ unit of B creates portfolio with terminal value of 20. 
\[\frac{1}{2} (10 ) + \frac{1}{2} (30) = 20\]
* Since we know prices of A and B, portfolio cost is 
\[-\frac{1}{2} (5) - \frac{1}{2} (15) = 10 \]
* Valuation by replication, therefore, says price of C is 10. 


# Illustration2

* Suppose price of C is 12. Is market in equilibrium?
``` {r myTable, echo=FALSE, message=FALSE, results='asis', comment=""}
if(!require(pander)){install.packages("pander");library(pander)}
ex_table = read.csv("./data/ex_table.csv")
names(ex_table) = c("","Time","")
pandoc.table(ex_table)
``` 


# Illustration2 : Solution
* No. Costless arbitrage profit is possible. Buy  $\frac{1}{2}$ unit of A, buy $\frac{1}{2}$ unit of B, and sell 1 unit of C
* Terminal value of overall portfolio equals 0.
\[\frac{1}{2} (10 ) + \frac{1}{2} (30) - 1 (20) = 0\]
* Cost of overall portfolio is 
\[-\frac{1}{2} (5) - \frac{1}{2} (15) + 1 (12) = 2 \]
* costless arbitrage profit is 2. 


# Valuation by replication
* No-arbitrage costless arbitrage opportunities (in a frictionless market) permits valuation by replication. 
  + Modigliani-Miller (1958, 1961) : Firm's dividend and capital structure decisions. 
  + Black-Scholes(1973) and Merton(1973) : Option valuation. 


# Rates of return for different holding periods
* Zero coupon bonds are bonds that sold at a discount from par value and provide their entire return from the difference between the purchase price and the ultimate repayment of par value
  + We can observe zero-coupon bonds with several different maturities
* Zero coupon bond, par = \$100, $T$ = maturity, $P$ = price, $r_f(T)$ = total risk free return
\[r_f(T) = \frac{100}{P(T)} -1\]
\end{itemize}
\end{frame}

# Annualized rates of return
figure

# Effective annual rate
* EAR : Effective annual rate
  + longer horizons provide greater total returns
  + We should compare the returns on investments with differing horizons
  + This comparison requires that we should express each total return as a rate of return for a common period
* EAR definition : percentage increase in funds invested over a 1-year horizon
\[1+EAR  = [1+r_f(T)]^{(1/T)}\]


# EAR: Example
* For the 6-month treasury , $T = 1/2$ and $r_f(0.5) = 2.71\%$
\[ 1 + EAR = (1 + 0.0271)^{(1/0.5)} \Rightarrow EAR = 5.49\%\]
* For the one-year treasury, $T = 1$ and $r_f(1) = 4.69\%$
\[ 1 + EAR = (1+ 0.0469)^1 \Rightarrow EAR = 4.69\%\]
* For the 25-year treasury, $T =25$ and $r_f(25) = 329.18 \%$
\[ 1+ EAR = (1+ 3.2918)^{1/25} \Rightarrow EAR = 6.00\% \] 


# Annual Percentage rates
* Rates on short-term investments often are annualized using simple rather than compound interest. These are called annual percentage rates (APRs)
\[  APR = \frac{r_f(T)}{T} \]
* APR : annualized using simple interst
  + For eample, the APR corresponding to a monthly rate is calculated by multiplying the monthly rate by 12
  + the APR of the 6-month bond is $2 \times 2.71\% = 5.42\%$

# APR vs. EAR
* The relationship among the compounding period, the EAR, and the APR is 
\[1+ EAR = (1+r_f(T))^{1/T} = (1+ T\times APR)^{1/T} \]
or 
\[APR = \frac{(1+EAR)^T -1 }{T}\]

* What is the limit of (1+EAR) as $T$ gets ever smaller? How far will these two rates diverge as the compounding frequency continue to grow? 
* We approach continuous compounding, and the relation of EAR to the annual percentage rate ($r_{cc}$) for the continuously compounded case is given by
\[1 + EAR = \lim_{T \rightarrow 0} (1+T\times APR)^{1/T} = e^{r_{cc}} \]
\item The total return for any period $T$ ($r_{cc}(T)$) is
\[ r_{cc}(T) = (1 + r_f(T)) = (1+EAR)^T = e^{ r_{cc} T} \]


# Interest Rate assumptions
* Assume interest rates are continuously compounded. 
* Virtue of using continuously compounded returns is correct time aggregation, i.e., T-day return equals sum of daily return
\[\ln\left(\frac{\tilde{S}_T}{S_0}\right) =\ln\left(\frac{\tilde{S}_1}{S_0}\right) +  \cdots + \ln\left(\frac{\tilde{S}_T}{\tilde{S}_{T-1}}\right) = \sum_{t=1}^{T} \ln\left(\frac{\tilde{S}_t}{\tilde{S}_{t-1}}\right) \]
\item T-day holding period returns
\[\frac{\tilde{S}_T - S_0}{S_0} \ne  \sum_{t=1}^{T} \left(\frac{\tilde{S}_t - \tilde{S}_{t-1}}{\tilde{S}_{t-1}}\right)\]


# Continuously Compounded Returns
* Assume asset price moves from 50 to 100 on first day, and then from 100 to 50 on next. 

|Day|Price|CCR|HPR|
|:---:|:---:|:---:|:---:|
|0| 50   |  | |
|1 |100  |  69.31\%|  100.00\% |
|2 |    50|     -69.31\% | -50.00\% |
|two-day return| |0.00\% | 0.00\% |
Sum(returns) | | 0.00\% | 50.00\% |


# Bond basics
* Notation
  + $r_t(t_1, t_2)$: interest rate from time $t_1$ to $t_2$ prevailing at time $t$
  + $P_{t_0}(t_1, t_2)$ : price of a bond quoted at $t = t_0$ to be purchased at $t=t_1$ maturing at $t = t_2$ 
  + Yield to maturity(YTM): percentage increase in dollars earned from th bond


# Bond basics
* Zero-coupon bonds
  + A bond that makes only a single payment at its maturity date. Zero-coupon bond price that pays $C_T$ at $T$ 
\[P(0,T) = \frac{C_T}{[1+r(0,T)]^T}\]
  + Yield curve : graph of annualized bond yields against time
    - A yield curve shows us how yields to maturity vary with time to maturity

# Implied forward rates
* Suppose current one-year rate $r(0,1)$ and two-year rate $r(0,2)$
  + \$1 investment for 1 year today : earns $1+r_0(0,1)$ 
  + \$1 investment for 2 years today : earns $[1+r_0(0,2)]^2$ 
  + What will be the implicit rate that can be earned from year 1 to year 2 ($r_0(1,2)$)
\begin{figure}
\includegraphics[width=0.6\textwidth]{fra}
\end{figure}
\end{itemize}
\end{frame}

# Bond basics
* zero-coupon bonds make a single payment at maturity 
 

# Discount Bonds
* Value of discount bond is
\[P_{t,T} = C_T e^{-r (T-t)} \]
* $P_{t,T}$ = price of bond at $t$ maturity $T$
* $C_T$ = face value of bond at maturity $T$
* $r$ = annualized zero-coupon (or spot) rate
* $T$ = term to maturity 

 
# Discount Bonds
* Value of discount bond is
\[P_{t,T} = C_T e^{-r (T-t)} \]
* Discount factor is today's price of one dollar received at time $T$
\[DF = e^{-r (T-t) }\]
* Implied interest rate is given current price is
\[r = \frac{\ln(C_T/P_{t,T})}{T-t}\]


# FRAs
* Consider a firm expecting to borrow \$100 million for 91 days, beginning 120 days from today. 
  + Borrows \$100 at time $t = 120$
  + Repay \$100 at time $t= 120 + 91 = 211$  
* Firm's risk? Increase in the cost of borrowing because the firm doesn't know the interest rate today
  + The 91-day interest rate ($r_{quarterly}$)  can be either 1.5\% or 2\% or maybe higher
  + How can we hedge this uncertainty?


# FRAs
* FRAs (Forward Rate Agreements) are over-the-counter contracts that guarantee a borrowing or lending rate on a given notional amount
* Can be settled at maturity (in arrears) or the initiation of the borrowing or lending transaction
* FRAs can be synthetically replicated using zero-coupon bonds


# FRAs
* FRAs are a forward contract based on the interest rate
  + The firm needs \$100 on day 120 and hopes to repay on 211 (90days loan)
  + Entering the FRAs is the same as the borrowing activity 
* Suppose the firm enters the FRAs with $r_{FRA} = 1.8\%$
  + FRA settlement in Arrears(Settle on 211) : $X = (r_{quarterly} - r_{FRA})\times notional$
  + At the time of borrowing (Settle on 120) : $X/(1+r_{quarter})$ 

 
# FRAs
* If the firm doesn't enter the FRAs
  + The borrowing cost : $r_{quarterly}$   :\textbf{Uncertain} 
* If the firm enters the FRAs
  + The borrowing cost : $A = - r_{quarterly}$
  + The FRAs : $B = (r_{quarterly} - r_{FRA})$
  + Ultimate borrowing cost : $A + B = -r_{FRA}$ : \textbf{Certain} 
* How to determine the fair FRAs in the market?
  + FRAs can be synthetically replicated using zero-coupon bonds


# FRAs
* Synthetic FRAs
  + $r_{FRA} = 1.8\%$ : Invest \$ 1 on day 120 and be guaranteed a 91-day return of 1.8\%

|Time | Day 0 | Day 120 | Day 211 |
|:---:|:---:|:---:|:---|
|Cash Flow | \$0 |  -\$1 | +\$1.018| 

  + To match the FRA cash flows

|Time| Day 0| Day 120 | Day 211|
|:---:|:---:|:---:|:---| 
|Borrow (Sell bonds)|  \$x|   -\$1 | 
|Lend (Buy bonds)|   -\$x |   +\$1.018|
|CF | \$0 | -\$1 | +\$1.018 |


# FRAs
* At time 0, we can observe the 120-day and the 211-day bonds traded on the market
  + $P(0,120) = 0.97561$
  + $P(0,211) = 0.95836$ 
* In order to receive \$1.018 on day 211, we buy 1.018 units of 211-day zero coupon bonds
  + The cost of this is $-1.018 \times 0.95836 = -0.97561$
  + At the same time, we sell 1 unit of 120-day bond and we receive $1 \times 0.97561$
  + This portfolio is equal to FRAs with ($r_{FRA} = 1.8\%$) 


# Forward rates of interest
* Forward rate on loan that begins at time $i$ and ends at time $j$ is determined by
\[e^{r_j T_j} = e^{r_i T_i} e^{f_{i,j} (T_j - T_i)} \]
* Re-arrange to isolate forward rate
\[f_{i,j} = \frac{r_j T_j - r_i T_i}{T_j - T_i}  \]


# Illustration
* Compute implied 91-day borrowing rate on loan beginning in 94 days. Assume you
  + Buy 94-day T-bills at ask rate : 5.123\%
  + Sell 185-day T-bills at bid rate : 5.177\%
\begin{eqnarray*}
T_i & = & 94/365 = 0.2575 \\
T_j & = & 185/365 = 0.5068 \\
f_{i,j} & = & \frac{0.05177 (0.5068) - 0.05123(0.2575)}{0.5068-0.2575} = 5.233\%
\end{eqnarray*}

# Forward rates
* How does loan work? Suppose we want to borrow 100 in 94 days. 
* Buy 94-day T-bills and sell 185-day T-bills.
\begin{small}
\begin{tabular}{lrrrrr}
      & Price & Unit & Today & 94 days & 185 days \\
\hline
Buy 94 TB  & 98.689 & 1.000 & -98.689 & 100.000 &  \\
Sell 185 TB & 97.410 & -1.013 & 98.689 &       & -101.313 \\
\cline{4-6}Net   &       &       & 0.000 & 100.000 & -101.313 \\
\end{tabular}
\end{small}
* Borrowing rate from 94days to 185 days is
\[\frac{\ln(101.313/100)}{91/365} = 5.233\% \]


# Summary
* Developing valuation and risk measurement principles requires assumptions about
investors and markets.
* Use two(but require only one):
  + law of one price : no costless arbitrage opportunities (required)
  + frictionless markets(made for convenience)

* No bid-ask spreads or commissions.
* No taxes.
* Unlimited borrowing or lending at risk-free rate of interest.
* Freedom to short sell with full use of proceeds.
* Can trade at any time and in any quantity. 