---
title: "07 Black Scholes Options Pricing"

---

# Purpose

Develop European-style option valuation formula

- The Black-Scholes formula is a limiting case of the binomial formula(infinitely many periods) for the price of a European option.
- Called "Black-Scholes" or "Black-Scholes/Merton" model

# Black-Scholes Assumptions

## Assumptions about stock return distribution

- Continuously compounded returns on the stock are normally distributed and independent over time, i.e. stock prices are log-normally distributed.
- The volatility of continuously compounded returns is known and constant
- Future dividends are known, either as dollar amount or as a fixed dividend yield

## Assumptions about the economic environment

- The risk-free rate is known and constant
- There are no transaction costs or taxes
- It is possible to short-sell costlessly and to borrow at the risk-free rate

# Black-Scholes Model

- Under risk-neutrality, current call value is
$$c = e^{-rT}E[\widetilde C_T]$$
- $E[\widetilde C_T]$ is the expected terminal call price based on asset price distribution
- Terminal call value is

\[
  \widetilde C_T = 
\left\{
\begin{aligned}
& \widetilde S_T - K, & if \quad S_T\ge K \\
& 0, & if \quad S_T \le K
\end{aligned}
\right.
\]

- Expected terminal call value of exercise price K is

\begin{equation} 
\begin{split}
E[\widetilde C_T] &= E[\widetilde S_T - K \mid S_T > K]Pr(S_T > K)\\
 & =E[\widetilde S_T \mid S_T > K]Pr(S_T > K) - E[K \mid S_T > K]Pr(S_T > K)\\
 & =E[\widetilde S_T \mid S_T > K] Pr(S_T > K) - KPr(S_T > K)
\end{split}
\end{equation} 

- Result does not depend on asset price distribution


# Expected value for Call option

- Under lognormality, expected value is

$$E[\widetilde C_T] = Se^{(\alpha - \delta)T}N(d_1) - KN(d_2)$$

$$where $$

\begin{equation} 
\begin{split}
d1 &= \frac{ln(\frac{S}{K}) + (\alpha - \delta + \frac{1}{2}\sigma^2)T}{\sigma \sqrt{T}}\\
d2 &= d1 - \sigma \sqrt{T}
\end{split}
\end{equation} 

## Standard Normal Distribution $N(d)$

- Use $z$ to represent a random variable that has a standard normal distribution

$$E[\widetilde Z] = 0 $$
$$Var(\widetilde Z) = 1$$

- The value of the cumulative normal distribution $N(d)$ equals to the probability $P(z < d)$ of a number $z$ drawn from the normal distribution to be less than $d$

$$N(d) = \int _{-\infty}^{d} \frac{1}{\sqrt{2 \pi}} e^-{\frac{1}{2}x^2dx}$$

- $N(- \infty) = 0$ and $N(\infty) = 1$
- Normal density is symmetric

$$N(-d) = 1 - N(d)$$

- $N(d)$ is cumulative normal probability. Area under curve to left of $d$
- In Excel, use **"=NORMSDIST(d)"**

```{r fig.cap='Cumulative Standard Normal'}
library(ggplot2)

x = seq(from = -3, to = 3, length.out = 100)
y = dnorm(x, mean  = 0, sd = 1)
df = data.frame(x, y)

ggplot(data = df, aes(x = x, y = y)) +
  geom_line() +
  xlab('Unit Normal Variable') +
  ylab('Probability') +
  geom_ribbon(
    data = subset(df, x <= 1.6),
    aes(ymax = y),
    ymin = 0,
    fill = "blue",
    colour = NA,
    alpha = 0.2
  ) +
  annotate("text", x = 0, y = 0.15, label = "N(d)") +
  geom_segment(data= df, mapping=aes(x=1.6, y=0.25, xend=1.6, yend=0),
               arrow=arrow(), size = 1, color="blue") +
  annotate("text", x = 1.6, y = 0.28, label = "d") +
  ggtitle('Cumulative Standard Normal') +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0, 0.1, 0)) 
```

## Call option price

- Take present value of expected future value to get European-style call formula

$$c = e^{-rT}E[\widetilde C_T]$$

- Thus,

$$c = Se^{(-r + \alpha - \delta)T}N(d_1) - Ke^{-rT}N(d_2)$$

$$where$$

\begin{equation} 
\begin{split}
d1 & = \frac{ln(\frac{S}{K}) + (\alpha - \delta + \frac{1}{2}\sigma^2)T}{\sigma \sqrt{T}} \\
d2 & = d1 - \sigma \sqrt{T}
\end{split}
\end{equation} 


- Black and Scholes(1973) value call on non-dividend paying stock
- Set $\alpha = r$ and $\delta = 0$

$$c = SN(d_1) - Ke^{-rT}N(d_2)$$

## Futures option valuation

- Black (1975) values call on futures
- Set $\alpha = r$ and $\delta = 0$

$$c = e^{−rT}[F N(d_1) − KN(d_2)]$$

$$where $$

\begin{equation} 
\begin{split}
d1 & = \frac{ln(\frac{F}{K}) + \frac{1}{2}\sigma^2T}{\sigma \sqrt{T}} \\
d2 & = d1 - \sigma \sqrt{T}
\end{split}
\end{equation} 


## Stock index option valuation

- Merton(1973) values call on index
- Set $\alpha = r$

$$c = Se^{-\delta T}N(d_1) - Ke^{-rT}N(d_2)$$

$$where$$

\begin{equation} 
\begin{split}
d1 & = \frac{ln(\frac{S}{K}) + (r - \delta + \frac{1}{2}\sigma^2)T}{\sigma \sqrt{T}} \\
d2 & = d1 - \sigma \sqrt{T}
\end{split}
\end{equation} 

## Currency option valuation

- Garman and Kohlhagen(1983) value call on foreign currency
- Set $\alpha = r_d$ and $\delta = r_f$

$$c = Se^{-r_f T}N(d_1) - Ke^{-r_dT}N(d_2)$$

$$where$$

\begin{equation} 
\begin{split}
d1 & = \frac{ln(\frac{S}{K}) + (r_d - r_f + \frac{1}{2}\sigma^2)T}{\sigma \sqrt{T}} \\
d2 & = d1 - \sigma \sqrt{T}
\end{split}
\end{equation} 


# Black-Scholes/Merton(BSM) Model

- BSM value of European-style call is

$$c = Se^{-\delta T}N(d_1) - Ke^{-rT}N(d_2)$$

$$where$$

\begin{equation} 
\begin{split}
d1 & = \frac{ln(\frac{S}{K}) + (r - \delta + \frac{1}{2}\sigma^2)T}{\sigma \sqrt{T}} \\
d2 & = d1 - \sigma \sqrt{T}
\end{split}
\end{equation} 

- Re-arrange expression for upper integral limit 


\begin{equation} 
\begin{split}
d1 & = \frac{ln(\frac{S}{K}) + (r - \delta + \frac{1}{2}\sigma^2)T}{\sigma \sqrt{T}}\\
 & = \frac{ln(\frac{Se^{-\delta T}}{Ke^{-rT}})+ \frac{1}{2}\sigma ^2 T}{\sigma \sqrt{T}}
\end{split}
\end{equation} 

- European-style put formula follows directly from
  - Put-call parity
  - European-style call formula
  
\begin{equation} 
\begin{split}
p & = Ke^{-rT} - Se^{-\delta T} + C \\
 & = Ke^{-rT} -Se^{-\delta T} + Se^{-\delta T}N(d_1) - Ke^{-rT}N(d_2) \\
 & = Ke^{-rT}[1-N(d_2)] -Se^{-\delta T}[1-N(d_1)] \\
 & = Ke^{-rT}N(-d_2) - S^{-\delta T}N(-d_1)
\end{split}
\end{equation}

- BSM value of European-style put is

$$P = Ke^{-rT}N(-d_2) - S^{-\delta T}N(-d_1)$$
$$where$$
\begin{equation} 
\begin{split}
d1 & = \frac{ln(\frac{Se^{-\delta T}}{Ke^{-rT}}) + \frac{1}{2}\sigma ^2 T}{\sigma \sqrt{T}} \\
d2 &= d1 - \sigma \sqrt{T}
\end{split}
\end{equation} 

## Illustration

- Value call and put options with $K = 40$ and $T = 3$ months
- Assume stock price is 41.00, stock pays no dividends, and stock’s volatility rate is 30% annually
- Assume interest rate is 8 % annually
- Download excel file
  - [Black Scholes Merton Option Pricing.xlsx](https://github.com/hyunyulhenry/bae_derivatives/raw/master/Excel/07%20Black%20Scholes%20Merton%20Option%20Pricing.xlsx)
  - Sheet : Option Valuation
  
- $Se^{-\delta T} \ and \ Ke^{-rT}$
$$Se^{-\delta T} = Se^{-0(3/12)} = 41$$
$$Ke^{-rT} = Ke^{-0.08(3/12)} = 39.21$$
- $N(d_1) \ and \ N(d2)$

\begin{equation} 
\begin{split}
d_1 & = \frac{ln(\frac{41}{39.21}) + 0.5(0.3^2)(0.25)}{0.3 \sqrt{0.25}} = 0.373 \\
d_2 & = 0.373 - 0.3 \sqrt{0.25} = 0.223
\end{split}
\end{equation} 

$$N(d_1) = N(0.373) = 0.6454$$
$$N(d_2) = N(0.223) = 0.5882$$

- Call option price

\begin{equation} 
\begin{split}
C & = SN(d_1) - Ke^{-rT}N(d_2) \\
  & = 41(0.6454) - 39.21(0.5882) \\
  & = 3.399
\end{split}
\end{equation}

- Put option price

\begin{equation} 
\begin{split}
P & = Ke^{-rT}N(-d_2) - SN(-d_1)  \\
  & = 39.21(1 - 0.5882) - 41(1 - 0.6454) \\
  & = 1.607
\end{split}
\end{equation} 

## Applying the model

- To use BSM formula to value options, we need estimates of
  - asset price : $S$ 
  - exercise price : $K$
  - time to expiration : $T$
  - income rate : $\delta$
  - interest rate : $r$
  - volatility : $\sigma$
  
- Some parameters are terms of option contracts
  - exercise price : $K$
  - time to expiration : $T$
  
- Some parameters are easily accessible market prices
  - asset price : $S$ $\rightarrow$ Use current bid/ask midpoint or last traded price
  - income rate : $\delta$ $\rightarrow$ Use dividend yield rate, foreign interest rate, lease rate etc
  - interest rate : $r$ $\rightarrow$ Use continuously compounded T-bill rate

- One parameter is more troublesome to estimate than others
  - volatility : $\sigma$
- Estimating volatility
  - Historical volatility
  - Implied volatility
  
# Volatility

- Volatility is unobservable
- Option prices can be quite sensitive to volatility
- One approach is to compute **historical volatility**: One problem with historical volatility is that expected future volatility can be different from historical volatility
- Alternatively, we can calculate **implied volatility**, which is the volatility that, when put into BSM formula, yields the observed option price

## Historical Volatility

- The logarithmic function computes daily returns from daily prices
$$r_{t+1} = ln(S_{t+1} / S_t)$$

- Since returns are additive, annual return is
$$r_{annual} = \sum_{t=1}^{252}r_t = \sum_{t=1}^{252}ln(S_t / S_{t-1}) = ln(S_{252} / S_0)$$

- The variance of the annual return is
$$Var(r_{annual}) = Var(\sum_{t=1}^{252}r_t)$$

- Suppose that returns are uncorrelated over time and that each day has the same variance of returns
$$Var(r_{annual}) = 252Var(r_t)$$

- Taking the square root of both sides yields
$$\sigma = \sqrt{252} \sigma_{daily}$$

- The term $\sqrt{252}$ should be modified depending the frequency of data
  - From monthly price data : $\sqrt{12}$
  - From weekly price data : $\sqrt{52}$

<div style="background-color:#F7F7F7;padding: 5px">
1. Take observations $S_0, S_1, \dots , S_T$ at interval of $\Delta t$ years. (e.g. for daily data $\Delta t = 1/252$)
2. Calculate the continuously compounded return in each interval as
$$r_t = ln(S_t/S_{t−1}) \ for \ t = 1, 2, \dots , T$$
3. Calculate the standard deviation, $\hat \sigma$, of the $r_t$’s
4. The annualized historical volatility is
$$\sigma = \frac{1}{\sqrt{\Delta t}} \hat \sigma$$
</div>

- Compute mean return from data. Use **"=AVERAGE()"**
$$\hat \mu = \frac{1}{T} \sum_{t=1}^{T}r_t$$

- Compute standard deviation. Use **"=STDEV()"**
$$\hat \sigma = \sqrt{\frac{1}{T-1}\sum_{t=1}^{T}(r_t - \hat \mu)^2}$$

- To annualize volatility, multiply by square root of number of calendar days(months, weeks, etc. ) in one year

### Historical Volatility-EWMA
- Computing volatility from standard deviation assumes the constant volatility through time
- Volatility appears to be changing over time
- If volatility is changing, we want to emphasize more recent observations
- One way to consider time-varying volatility is to compute an *Exponentially Weighted Moving Average(EWMA)* of the squared stock returns 

- Equal weighted squared returns
$$\hat \sigma = \sqrt{\frac{1}{T} \sum_{t=1}^{T} r_{t}^{2}} $$

- Exponentially weighted moving average volatility
$$\hat \sigma_t = \sqrt{\sum_{i=1}^{n} \left[\frac{(1 - \lambda)\lambda^{i-1}}{\sum_{j=1}^{n}(1 - \lambda)\lambda^{j-1}} \right]r_{t-i}^{2}}$$
  - The lower is the value of λ, the higher is the weight placed on the most recent data

- Compute historical volatilities for AAPL
- Download five-year daily price data from http://finance.yahoo.com
- Use the **adjusted price** series which are close prices adjusted for dividends and stock-splits
  - If there are no adjusted prices, you should adjust observed closed prices for cash dividends and stock splits after getting the distribution information
- Based on adjusted close prices, compute continuously compounded returns
- Download excel file
  - [Historical Volatility.xlsx](https://github.com/hyunyulhenry/bae_derivatives/raw/master/Excel/07%20Historical%20Volatility.xlsx)
  - Sheet : Historical Volatility

```{r}
library(magrittr)
library(knitr)
library(kableExtra)

data.frame(
    V1 = c('All Date', 'One year', '90 days', '30 days'),
    V2 = c('1,257', '251', '90', '30'),
    V3 = c('1.69%', '1.40%', '1.26%', '0.95%'),
    V4 = c('26.85%', '22.20%', '20.04%', '15.05%')
  ) %>% kable(booktabs = T, escape = F, align = "r",
      caption = "Estimates depend on length of return history",
      col.names = c('Data Length', 'N', 'StDev (Daily)', 'StDev (Annualized)')) %>%
  kableExtra::kable_styling(latex_options = c("striped", "hold_position")) 
```

- Download excel file
  - Historical Volatility.xlsx
  - Sheet : EWMA
  
```{r}
data.frame(
    V1 = c('0.25', '0.50', '0.75'),
    V2 = c('0.35%', '0.51%', '0.64%'),
    V3 = c('5.56%', '8.11%', '10.20%')
  ) %>% kable(booktabs = T, escape = F, align = "r",
      caption = "Volatility of Apple returns appeared to be lower later
in the year",
col.names = c('$\\lambda$', 'EWMA (Daily)', 'EWMA (Annualized)')) %>%
  kableExtra::kable_styling(latex_options = c("striped", "hold_position")) 
```  

## Implied Volatility

- Implied volatility is computed from observed option prices
- Set option prices equal to formula value and solve for $\sigma$

$$c = f(S, K, T, r, \delta, \sigma)$$

  - $c$ : observed call option price
  - $f()$: BSM formula
  - $S$ : observed stock price
  - $K$  and T : contract parameters
  - $r$ : risk-free rate
  - $\delta$ : dividend yield
  
- [Download AAPL option prices from CBOE website](http://www.cboe.com/DelayedQuote/QuoteTableDownload.aspx)
- Download excel file
  - [Implied Volatility.xlsx](https://github.com/hyunyulhenry/bae_derivatives/raw/master/Excel/07%20Implied%20Volatility.xlsx)
  - Sheet : IV
  
- Some practical uses of implied volatility include
  - Use the implied volatility from an option with an observable price to calculate the price of another option on the same underlying asset
  - Use implied volatility as a quick way to describe the level of options prices on a given underlying asset: you could quote option prices in terms of volatility, rather than as a dollar price
  - Checking the uniformity of implied volatilities across various options on the same underlying assets allows one to verify the validity of the pricing model in pricing those options

# Summary

<div style="background-color:#F7F7F7;padding: 5px">

- Use risk-neutrality and lognormal asset price distribution to value European-style options
  - Black-Scholes stock option formula
  - Black futures option formula
  - Merton constant dividend yield option formula
  - Garman/Kohlhagen currency option formula
  
- The contribution of Black-Scholes was not the particular formula but rather the appearance of the risk-free rate in the formula

- Applying option valuation models require estimates of
  - exercise price($K$) and time to expiration($T$)
  - underlying asset price($S$)
  - income rate ($\delta$)
  - interest rate ($r$)
  - volatility($\sigma$)
    - Historical volatility
    - Implied valatility
    
</div>